Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-06-06 paper.Rnw
edits through sec 3

We perform six applications of LIME with different tuning parameters to the random forest predictions from all observations in the sine data test set. To do this, we use the R package \emph{limeaid} \footnote{Available on GitHub at https://github.com/goodekat/limeaid} (\Sexpr{packageVersion("limeaid")}), which applies a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (\Sexpr{packageVersion("lime")}) \citep{pedersen:2020}. The forked version allows us to export internal values from \emph{lime}. The tuning parameters are as follows. For each application, we specify that LIME returns the top two features in the explanation, because we expect $x_1$ and $x_2$ to be the important features. Note that the complexity of an explanation goes hand in hand with the number of features selected to be included in the explanation. Setting the number of features to a small number will make the explanation simpler but not necessarily more correct. A quantile bin based simulation method is used for five of the implementations with the number of bins varying from 2 to 6 by application. The sixth implementation of LIME uses a kernel density simulation method. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications.

@@

# Determine the bin cuts for the default lime implementation
sine_lime_default_bin_cuts <- sine_lime_default %>%
  select(feature, feature_desc) %>%
  unique() %>%
  separate(feature_desc, c("other", "bin_cut"), sep = " <= ") %>%
  select(-other) %>%
  na.omit() %>%
  mutate(bin_cut = as.numeric(bin_cut)) %>%
  arrange(feature) %>%
  mutate(case = rep(1:3, 2)) %>%
  spread(feature, bin_cut)
@

@@

The \data \ explanation only includes two features. In situations were more than two features are included in a LIME explanation, the explanation scatterplots can be extended to a generalized scatterplot matrix \kgc{(reference)} that includes all pairwise combinations of features. Generalized scatterplot matrices (and scatterplot matrices in general) fail to provide helpful information when the number of features becomes large \kgc{(reference)}. It is common for machine learning models to have a large number of features, and therefore, a generalized scatterplot matrix of explanation scatterplots for all features would be ineffective. However, in the application of LIME, the user selects the number of features to return in the explanation (as previously mentioned). \citet{pedersen:2020} encourage users to select less than 10 features to include in the explanation\footnote{https://cran.r-project.org/web/packages/lime/vignettes/Understanding\_lime.html}. As a result, specifying a small number of features to include in a LIME explanation makes it feasible to use generalized scatterplot matrices of explanation scatterplots.

@@

The feature heatmap uses colors to identify the features selected by LIME across tuning parameters for sets of explanations (referred to as cases here) faceted by the position of feature importance assigned by LIME. That is, for LIME applied with $t$ sets of tuning parameters to $n$ cases to select the $f$ top features, create $f$ heatmaps (one for each of the positions of importance determined by the magnitude of the explainer model coefficients) with the cases on the y-axis, the tuning parameters on the x-axis, and the cells colored by the feature chosen for the corresponding case and tuning parameter. This plot is used assess the locality of the explanations and to compare results from different input options.

@@

\autoref{fig:sine-feat-heat} shows the feature heatmap for the LIME applications to the 100 observations in the \data \ test set. The most important (top facet) and second most important (bottom facet) features are shown in this plot. For the quantile bins, the original features prior to the indicator variable transformation are included since it is obvious that different features would be selected when the sizes of the bins change. This figure shows that for the kernel density and 2 quantile bin implementations, LIME selects $x_1$ as the most important feature and $x_2$ as the second most important feature across all cases in the test set. These explanations are not local. There is variability in the features selected by LIME for 3 to 6 quantile bins. For these tuning parameters, there is a mix of horizontal and vertical stripes, which suggests a dependence on tuning parameters. Note that the explanations from 3 and 5 quantile bins include the selection of the random noise variable ($x_3$) as an important variable in many predictions, which should not be the case. The pattern seen in the explanations for 3 to 6 quantile bins may suggest local explanations, but the dependence on tuning parameter makes it unclear which set of explanations to use.

@@

@
\caption{Example of a metric comparison plot using the \data. Each row of the plot corresponds to one of three metrics: average fidelity, average $R^2$, or MSEE. The LIME tuning parameters are plotted on the x-axis, and the metric values are shown on the y-axis. The points are colored by rank within a metric indicating best to worst (dark to light). The kernel density simulation methods performs well according to all three metrics.}
\label{fig:sine-assess-metric}
\end{figure}

@@

Each metric presented below is computed on a set of LIME explanations associated with a set predictions of interest obtained using the the same tuning parameters. Here we provide a high level description of the metrics. Notation and formulas for these metrics are included in Appendix \ref{metric-details}.

\begin{itemize}
\item \emph{Average $R^2$}: A metric for the assessment of the model fit and linearity assumption computed as the average of ridge regression explainer models $R^2$ values (deviance ratios from the R package \emph{glmnet}).

\item \emph{Average Fidelity}: Provides comparison of the complex and simple model predictions for each of the values in the simulated dataset accounting for the assigned weight computed as the average of the fidelity metric presented in \citet{ribeiro:2016} (weighted distance between explainer and complex model predictions for simulated data).

\item \emph{Mean Squared Explanation Error (MSEE)}: Measures the faithfulness of the explainer model to the complex model by comparing their predictions similar to the average fidelity, but only the observation of interest is used to compute the average squared deviation over the observations in the test set.
\end{itemize}

\autoref{fig:sine-assess-metric} shows these metrics computed for each of the different tuning parameter applications of LIME to the \data. The simulation methods are listed on the x-axis. The plot is faceted by metric, and the metric values are plotted on the y-axis. The colors of the points represent the rank of the simulation methods performance based on a particular metric (darker indicates a better metric value and lighter indicates a worse metric value). Higher average $R^2$ values are better, and lower average fidelity and MSEE values are better.

All three metrics suggest that the kernel density implementation performed well, but the metrics disagree for the quantile bins methods. Average $R^2$ values and average fidelity rank the performance of the number of quantile bins the same (2 quantile bins perform the best and 6 quantile bins perform the worst). In fact, these two metrics appear to have an inverse relationship in this example. MSEE provides different ranks of the simulation methods. MSEE suggests that 6 quantile bins perform the best, and 2 quantile bins perform the worst. This is the opposite of average $R^2$ and average fidelity. Since average fidelity and MSEE are similar metrics, it would be expected that they would agree, but that is so in this example. This may be due to the MSEE only taking into account the prediction of interest and not the full simulated dataset. The contradiction between metrics makes it difficult to identify which simulation method explanations to trust.

Recall that \autoref{fig:sine-feature-heatmap} indicated that the LIME kernel density method selected the same feature across all cases in the test dataset for both the first and second features. It appears that a global trend may be the best explanation for this example, which may be reasonable considering that we know that both $x_1$ and $x_2$ are the two features that should be the features used by the random forest to distinguish between response categories. At this point, the use of explanation scatterplots for individual predictions of interest may provide better insight into a selection of tuning parameters to use.

@@

The \data \ is a relatively simple problem where the true decision boundary can be easily visualized and diagnostics for the performance of LIME can directly be compared to ground truth. In practice, we usually encounter more complicated situations. The remainder of this section provides a discussion of the visual diagnostics at the example of a practical data problem investigating the similarity of marks on fired bullets.

@@

\begin{figure}[!thp]
\centering
\includegraphics[width=2.5in]{./figure-static/bullets.png}
\caption{(Top left) Traditionally rifled gun barrel. Grooves and lands alternate to give bullets a spin during the firing process.
Barrel create markings (striations) on a bullet when fired.
(Top right) Image of a fired bullet. The vertical stripes along the lower half of the bullet show groove and land engraved areas. The land engraved areas contain the microscopic striations created when the bullet passed through the barrel of the gun. (Bottom) Zoomed in image of a land engraved area showing striations (vertical lines).}
\label{fig:bullet}

@@

Since firearm identification is commonly used as evidence for convictions in court cases, it is important to be able to understand and assess the model that is used to quantify the probability that a bullet was fired from a gun. An application of LIME to rtrees could provide an understanding of the key variables used by the random forest model to make a prediction. However, just as it is important to assess the random forest model for this high-stakes application, it is also important to assess the LIME explanations to make sure they are providing a trustworthiness understanding of the random forest model. We will apply LIME to rtrees and use our visual diagnostics to assess different implementations of LIME.

@@

# Determine the implementation and case number of the poi
# in the no_perms data
poi_cases_no_perms <- bullet_explain_noperms %>%
  filter(case %in% c(bullet_poi_match, bullet_poi_nonmatch),
         nbins %in% c(3, NA),
         gower_pow == 0.5) %>%
  select(case, implementation) %>%
  unique()

@@

@
\caption{Feature heatmap of the 36 LIME implementations applied to the bullet comparison data test set. In addition to faceting the results by simulation method and order a feature was selected by LIME as done in \autoref{fig:sine-feature-heatmap}, this plot has been faceted by the Gower power and whether the the observation is from a match (TRUE) or non-match (FALSE). This plot shows several key insights about the LIME explanations. First, the features selected by LIME for the density simulation methods are mostly the same for all implementations and all cases. Second, the features selected by LIME for the bin based methods are affected by the number of bins but not the Gower power. Third, the explanations produced by LIME using bin based methods result in different important features for the matches and non-matches. \kgc{***Need to rotate the facet labels, so I can make the text bigger.}}
\label{fig:bullet-feature-heatmap}
\end{figure*}

We apply LIME to each prediction from the bullet test data obtained from the 'rtrees' random forest model for each of the four sampling methods (equally spaced bins, quantile bins, kernel density estimation, and normal approximation). For each of the bin based sampling methods, LIME was applied for 2 to 6 bins. It was decided to use 6 bins as the maximum, since the larger the number of bins, the more complex the explanation becomes. Each simulation method was implemented three times with Gower exponents of 0.5, 1, and 10. Thus, a total of $12\times 3=36$ different implementations of LIME were performed. These implementations of LIME were performed using our R package \emph{limeaid}. Each implementation was set to return 3 features by LIME and feature selection was performed using the highest weights method.

@@

To get an overview of the LIME explanations from the 36 implementations, we consider our heatmap diagnostic plot for comparing the LIME implementations first shown in \autoref{fig:bullet-feature-heatmap}. The plot facets the features selected in the LIME explanations by the simulation method, the Gower power, the order the feature was chosen, and whether the observation is from a known match or non-match of bullet signatures. This plot highlights several key features of the LIME explanations from the bullet matching dataset.

First of all, the density simulation methods produced the same explanations for almost all cases and LIME implementation methods. As a result, the LIME explanations for the density implementation methods are global and not local. Second, within a bin based simulation method, the features selected by LIME for an observation often vary based on the number of bins used but do not appear to vary based on the Gower power used. Especially with the equal bin explanations, there are vertical stripes that suggest a dependence of the LIME explanations on the number of bins used. The vertical stripes are not as apparent with the quantile bins. Lastly, there are clear differences between the LIME explanations for the cases that are matches and those that are not matches which are produced by the bin based simulation methods. This provides evidence that suggests that the features which the random forests uses to classify a match are different from the features used by the random forest to classify a non-match. The most important insight \autoref{fig:bullet-feature-heatmap} highlights is the dependence of the explanations on the LIME implementation method for many of the observations in the test data.

@@

@
\caption{Plot of LIME assessment metrics for the 36 different implementations of LIME applied to the bullet comparison data test set. The shape of the points represents the power used for the Gower distance calculation, and the color of the points represents the rank associated with that observation within a metric. The density simulation methods perform well for all metrics, but the bin based metrics often do not agree in terms of performance across metrics.}
\label{fig:bullet-assess-metric}
\end{figure*}

To try to identify the LIME implementation method with the most trustworthy set of explanations, we created an assessment metric plot shown in \autoref{fig:bullet-assess-metric}. The three metrics of average fidelity, average $R^2$, and MSEE were computed for each of the simulation methods and Gower power used to implement LIME. The plot is faceted by the simulation method, and the shape of a point represents the Gower power. The color of a point represents the rank within a metric. That is, the darker a point, the better performance of that implementation method based on the metric.

The density simulation methods performed well across all implementation methods. This is an interesting result, since \autoref{fig:bullet-feature-heatmap} shows that the density methods results in global explanations. The metric values for the bin based implementation methods do not agree across metrics. For example, 2 and 3 quantile bins performed well according to the average $R^2$ but poorly according to MSEE. As a result, with the bin based methods, it is difficult to know which method to recommend. When considering the metrics across the Gower power used, the implementations using a power of 0.5 perform best or as well as the other powers in across all simulation methods. This suggests that the power that leads to a more global explanation is preferred by LIME in this example.

@@

Figures \ref{fig:bullet-eois-nonmatch} and \ref{fig:bullet-eois-match} contain plots that provide this closer look for the known non-match and match, respectively. The top row of plots in each figure are the explanation plots from \emph{lime}. The bottom row of plots show scatterplots of the features selected by LIME colored by the random forest model prediction. The location of the prediction of interest is indicated on the scatterplots by a diamond, and black solid lines are included on the plots representing the bin divisions for implementations using bins. The bottom row plots are created using \emph{limeaid}. The plots within a column are associated with one of the three implementation methods (kernel density, 3 equal bins, or 3 quantile bins).

The scatterplots provide a visualization that helps to explain the LIME explanation. For example, consider the explanation for case \Sexpr{bullet_poi_nonmatch} when 3 quantile bins were used. The explanation plot from \emph{lime} shows that case \Sexpr{bullet_poi_nonmatch} having an observed CCF value greater than 0.320 supports a random forest prediction in favor of a match. The scatter plots show that many of the simulated values with CCF greater then 0.320 have random forest predictions greater than 0.5. Additionally, the LIME explanation indicates that a value of matches less than or equal to 1.66 contradicts a prediction in favor of a match. The scatter plots shows that almost all simulated values with a value of matches less than or equal to 1.66 have random forest predictions close to 0. Similar statements can be made about the 3 equal bin explanation for this case and the bin based implementations for the known match observation explanations.

@@

\begin{figure*}[!thp]
<<bullet-eois-nonmatch, out.width = '\\textwidth', fig.width = 12, fig.height = 6.25, message = FALSE>>=
# Join the plots
plot_grid(pf2_3qb, pf2_3eb, pf2_kd,
          eoi2_3qb, eoi2_3eb, eoi2_kd,
          nrow = 2,
          rel_heights = c(0.45, 0.55))
@
\caption{Plots of LIME explanations for one case in the test data that is a known non-match. Each column contains plots associated with a different implementation (3 quantile bins, 3 equal bins, or kernel density). The top row of plots are the explanation plots from \emph{lime}. The bottom row shows scatter plots of the simulated data associated with the prediction of interest. The features plotted are those chosen by LIME in the explanations, and the points are colored by the the random forest predictions. Lines showing the divisions created in the feature space by the bin based LIME methods are included as solid black lines. The scatter plots can be used to better understand the LIME explanation and assess if it is a good explanation.}
\label{fig:bullet-eois-nonmatch}

@@

The scatter plots of the kernel density implementations for both the match and non-match case show that almost all of the simulated values have random forest predictions less than 0.5, which provide evidence in favor of a non-match. It appears that simulation method is not providing enough values to cover the feature space that hit regions where the random forest provides predictions above 0.5. The training data has a much smaller amount of matches (\Sexpr{sum(bullet_train$samesource)}) than non-matches (\Sexpr{sum(bullet_train$samesource == FALSE)}), and this simulation method is clearly affected by this imbalance in the two classification categories. Furthermore, this may be the cause of the minimal amount of variability in the LIME explanations across all of the cases in the test data.

Without applying multiple implementations LIME to the bullet test data or viewing diagnostic plots of the LIME explanations, it may be very possible to formulate reasons why the LIME explanations make sense. However, the sequence of plots in this section (Figures \ref{fig:bullet-feature-heatmap}, \ref{fig:bullet-lime-expls}, \ref{fig:bullet-assess-metric}, \ref{fig:bullet-eois-match}, and \ref{fig:bullet-eois-nonmatch}) suggest that we should be cautious to trust any of these LIME explanations. While some of them provide an explanation that may be reasonable, there could be better explanations (as seen in the 3 quantile bin case of \autoref{fig:bullet-eois-match}), and some of them provide explanations that make no sense (as seen in \autoref{fig:bullet-lime-expls}). It appears that either LIME needs to be further tuned to provide trustworthy and good explanations, or a different approach may provide better insight. As glimpsed with the scatter plots in Figures \ref{fig:bullet-eois-match} and \ref{fig:bullet-eois-nonmatch}, the more traditional approach to explanation random forest predictions by plotting the random forest predictions over the feature space (partial dependence plots), could provide better insight into the complex model with no need to assess the explanation method if the sampling method is well chosen.

@@

The figures referenced in the previous paragraph from the \data \ example are all based on the implementation of LIME using a set of tuning parameters. From the use of the feature heatmap, we found that different sets of tuning parameters can produce different explanations. It makes sense that one set of tuning parameters could lead to a more faithful explainer model than another. As a result, we proposed a visual comparison of two faithfulness metrics (MSEE and average fidelity). The examples of faithfulness metric comparisons in this paper (Figures \ref{fig:sine-assess-metric} and \ref{fig:bullet-assess-metric}) both produced confusing results. In both examples, the density based simulation methods resulted in the best or close to the best performance even though the feature heatmaps (Figures \ref{fig:sine-feature-heatmap} and \ref{fig:bullet-feature-heatmap}) showed that the density simulation methods produced global explanations with no variation in features selected by LIME. For the bin based simulation methods, the two metrics often did not agree or contradicted one another, which makes it difficult to decide on a recommendation of a set of tuning parameters that produces the explanations with the most faithful explainer model.

>>>>

We perform six applications of LIME with different tuning parameters to the random forest predictions from all observations in the sine data test set. To do this, we use the R package \emph{limeaid} \footnote{Available on GitHub at https://github.com/goodekat/limeaid} (\Sexpr{packageVersion("limeaid")}), which applies a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (\Sexpr{packageVersion("lime")}) \citep{pedersen:2020}. The forked version allows us to export internal values from \emph{lime}. The tuning parameters are as follows. For each application, we specify that LIME returns the top two features in the explanation, because we expect $x_1$ and $x_2$ to be the important features. Note that the complexity of an explanation goes hand in hand with the number of features selected to be included in the explanation. Setting the number of features to a small number will make the explanation simpler but not necessarily more correct. A quantile bin based simulation method is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. The sixth application of LIME uses a kernel density simulation method. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications.

@@

# Determine the bin cuts for the default lime application
sine_lime_default_bin_cuts <- sine_lime_default %>%
  select(feature, feature_desc) %>%
  unique() %>%
  separate(feature_desc, c("other", "bin_cut"), sep = " <= ") %>%
  select(-other) %>%
  na.omit() %>%
  mutate(bin_cut = as.numeric(bin_cut)) %>%
  arrange(feature) %>%
  mutate(case = rep(1:3, 2)) %>%
  spread(feature, bin_cut)
@

@@

The \data \ explanation only includes two features. In situations were more than two features are included in a LIME explanation, the explanation scatterplots can be extended to a generalized pairs plot \citep{emerson:2013} that includes all pairwise combinations of features. Generalized pairs plots (and scatterplot matrices in general) fail to provide helpful information when the number of features becomes large. It is common for machine learning models to have a large number of features, and therefore, a generalized pairs plot of explanation scatterplots for all features would be ineffective. However, when applying LIME, the user selects the number of features to return in the explanation (as previously mentioned). In the \emph{lime} R package, \citet{pedersen:2020} encourage users to select less than 10 features to include in the explanation. As a result, specifying a small number of features to include in a LIME explanation makes it feasible to use generalized pairs plot of explanation scatterplots.

@@

The feature heatmap uses colors to identify the features selected by LIME across tuning parameters for sets of explanations (referred to as cases here) faceted by the position of feature importance assigned by LIME. That is, for LIME applied with $t$ sets of tuning parameters to $n$ cases to select the $f$ top features, create $f$ heatmaps (one for each of the positions of importance determined by the magnitude of the explainer model coefficients) with the cases on the y-axis, the tuning parameters on the x-axis, and the cells colored by the feature chosen for the corresponding case and tuning parameter. Additional tuning parameters may also be included in the plot via facets. This plot is used assess the locality of the explanations and to compare results from different input options.

@@

\autoref{fig:sine-feat-heat} shows the feature heatmap for the LIME applications to the 100 observations in the \data \ test set. The most important (top facet) and second most important (bottom facet) features are shown in this plot. For the quantile bins, the original features prior to the indicator variable transformation are included since it is obvious that different features would be selected when the sizes of the bins change. This figure shows that for kernel density and 2 quantile bins, LIME selects $x_1$ as the most important feature and $x_2$ as the second most important feature across all cases in the test set. These explanations are not local. There is variability in the features selected by LIME for 3 to 6 quantile bins. For these tuning parameters, there is a mix of horizontal and vertical stripes, which suggests a dependence on tuning parameters. Note that the explanations from 3 and 5 quantile bins include the selection of the random noise variable ($x_3$) as an important variable in many predictions, which should not be the case. The pattern seen in the explanations for 3 to 6 quantile bins may suggest local explanations, but the dependence on tuning parameter makes it unclear which set of explanations to use.

@@

@
\caption{\emph{Assessment metric plot}. The assessment metrics computed on the applications of LIME to the \data \ test set are included in this example assessment metric plot. Each horizontal facet corresponds to one of three metrics: average $R^2$, average fidelity, or MSEE. The LIME tuning parameter (simulation methods in this example) is plotted on the x-axis, and the metric values are shown on the y-axis. The points are colored by rank (within a metric) indicating best to worst (dark to light). The kernel density simulation methods performs well according to all three metrics.}
\label{fig:sine-assess-metric}
\end{figure}

@@

Each metric presented below is computed on a set of LIME explanations obtained using the the same tuning parameters. Here we provide a high level description of the metrics. Notation and formulas for these metrics are included in Appendix \ref{metric-details}.

\begin{itemize}
\item \emph{Average $R^2$}: Assess the model fit and linearity claim by computing the average of the explainer model $R^2$ values (deviance ratios from the R package \emph{glmnet}).

\item \emph{Average Fidelity}: Measures the faithfulness of the explainer model to the complex model by comparing their predictions. Computed as the average of the explainer model fidelity metrics: a metric presented in \citet{ribeiro:2016} (the weighted distance between explainer and complex model predictions for all observations in the LIME simulated data associated with an individual prediction of interest).

\item \emph{Mean Squared Explanation Error (MSEE)}: Also measures the faithfulness of the explainer model to the complex model by comparing their predictions, but only the prediction of interest is used to compute an average squared deviation between explainer and complex model predictions.
\end{itemize}

\autoref{fig:sine-assess-metric} shows these metrics computed for each of the LIME applications to the \data \ test set. The simulation methods are listed on the x-axis. The plot is faceted by metric, and the metric values are plotted on the y-axis. The colors of the points represent the rank of the simulation methods performance based on a particular metric (darker indicates a better metric value and lighter indicates a worse metric value). Higher average $R^2$ values are better, and lower average fidelity and MSEE values are better. This example only includes one tuning parameter: the simulation method. If more than one tuning parameter is considered, the assessment metric plot is extended by adding additional facets or levels to the x-axis.

All three metrics suggest that the kernel density performed well, but the metrics disagree for the quantile bins methods. Average $R^2$ and average fidelity rank the performance of the number of quantile bins the same (2 quantile bins perform the best and 6 quantile bins perform the worst). In fact, these two metrics appear to have an inverse relationship in this example. MSEE provides different ranks of the simulation methods. MSEE suggests that 6 quantile bins perform the best, and 2 quantile bins perform the worst. This is the opposite of average $R^2$ and average fidelity. Since average fidelity and MSEE are similar metrics, it would be expected that they would agree, but that is so in this example. This may be due to the MSEE only taking into account the prediction of interest and not the full simulated dataset. The contradiction between metrics makes it difficult to identify which simulation method explanations to trust.

Recall that \autoref{fig:sine-feat-heat} indicated that the LIME kernel density method selected the same feature across all cases in the test set for both the first and second features. It appears that a global trend may be the best explanation for this example, which may be reasonable considering that we know that both $x_1$ and $x_2$ are the two features that should be the features used by the random forest to distinguish between response categories.

@@

In this section, we provide a discussion of the application of the visual diagnostics for LIME explanations to a practical data problem investigating the similarity of marks on fired bullets.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{./figure-static/bullets.png}
\caption{(Top left) Traditionally rifled gun barrel. Grooves and lands alternate to give bullets a spin during the firing process.
Barrel create markings (striations) on a bullet when fired.
(Top right) Image of a fired bullet. The vertical stripes along the lower half of the bullet show groove and land engraved areas. The land engraved areas contain the microscopic striations created when the bullet passed through the barrel of the gun. (Bottom) Zoomed in image of a land engraved area showing striations (vertical lines).}
\label{fig:bullet}

@@

Since firearm identification is commonly used as evidence for convictions in court cases, it is important to be able to understand and assess the model that is used to quantify the probability that a bullet was fired from a gun. An application of LIME to rtrees could provide an understanding of the key variables used by the random forest model to make a prediction. However, just as it is important to assess the random forest model for this high-stakes application, it is also important to assess the LIME explanations to make sure they are providing a trustworthiness understanding of the random forest model. We will apply LIME to rtrees and use our visual diagnostics to assess different LIME tuning parameters.

@@

# Determine the application and case number of the poi
# in the no_perms data
poi_cases_no_perms <- bullet_explain_noperms %>%
  filter(case %in% c(bullet_poi_match, bullet_poi_nonmatch),
         nbins %in% c(3, NA),
         gower_pow == 0.5) %>%
  select(case, implementation) %>%
  unique()

@@

@
\caption{Feature heatmap of the 36 LIME applications to the bullet comparison data test set. In addition to faceting the results by simulation method and order a feature was selected by LIME as done in \autoref{fig:sine-feature-heatmap}, this plot has been faceted by the Gower power and whether the the observation is from a match (TRUE) or non-match (FALSE). This plot shows several key insights about the LIME explanations. First, the features selected by LIME for the density simulation methods are mostly the same for all tuning parameters and all cases. Second, the features selected by LIME for the bin based methods are affected by the number of bins but not the Gower power. Third, the explanations produced by LIME using bin based methods result in different important features for the matches and non-matches. \kgc{***Need to rotate the facet labels, so I can make the text bigger.}}
\label{fig:bullet-feature-heatmap}
\end{figure*}

We apply LIME to each prediction from the bullet test data obtained from the 'rtrees' random forest model for each of the four sampling methods (equally spaced bins, quantile bins, kernel density estimation, and normal approximation). For each of the bin based sampling methods, LIME was applied for 2 to 6 bins. It was decided to use 6 bins as the maximum, since the larger the number of bins, the more complex the explanation becomes. Each simulation method was implemented three times with Gower exponents of 0.5, 1, and 10. Thus, a total of $12\times 3=36$ different applications of LIME were performed. These applications of LIME were performed using the R package \emph{limeaid}. Each application was set to return 3 features by LIME and feature selection was performed using the highest weights method.

@@

To get an overview of the LIME explanations from the 36 applications, we consider our heatmap diagnostic plot for comparing the LIME applications first shown in \autoref{fig:bullet-feature-heatmap}. The plot facets the features selected in the LIME explanations by the simulation method, the Gower power, the order the feature was chosen, and whether the observation is from a known match or non-match of bullet signatures. This plot highlights several key features of the LIME explanations from the bullet matching dataset.

First of all, the density simulation methods produced the same explanations for almost all cases and LIME tuning parameters. As a result, the LIME explanations for the density methods are global and not local. Second, within a bin based simulation method, the features selected by LIME for an observation often vary based on the number of bins used but do not appear to vary based on the Gower power used. Especially with the equal bin explanations, there are vertical stripes that suggest a dependence of the LIME explanations on the number of bins used. The vertical stripes are not as apparent with the quantile bins. Lastly, there are clear differences between the LIME explanations for the cases that are matches and those that are not matches which are produced by the bin based simulation methods. This provides evidence that suggests that the features which the random forests uses to classify a match are different from the features used by the random forest to classify a non-match. The most important insight \autoref{fig:bullet-feature-heatmap} highlights is the dependence of the explanations on the LIME tuning parameter for many of the observations in the test data.

@@

@
\caption{Plot of LIME assessment metrics for the 36 different applications of LIME applied to the bullet comparison data test set. The shape of the points represents the power used for the Gower distance calculation, and the color of the points represents the rank associated with that observation within a metric. The density simulation methods perform well for all metrics, but the bin based metrics often do not agree in terms of performance across metrics.}
\label{fig:bullet-assess-metric}
\end{figure*}

To try to identify the set of LIME tuning parameters with the most trustworthy set of explanations, we created an assessment metric plot shown in \autoref{fig:bullet-assess-metric}. The three metrics of average fidelity, average $R^2$, and MSEE were computed for each of the simulation methods and Gower power used to implement LIME. The plot is faceted by the simulation method, and the shape of a point represents the Gower power. The color of a point represents the rank within a metric. That is, the darker a point, the better performance of that method based on the metric.

The density simulation methods performed well across all tuning parameters. This is an interesting result, since \autoref{fig:bullet-feature-heatmap} shows that the density methods results in global explanations. The metric values for the bin based methods do not agree across metrics. For example, 2 and 3 quantile bins performed well according to the average $R^2$ but poorly according to MSEE. As a result, with the bin based methods, it is difficult to know which method to recommend. When considering the metrics across the Gower power used, the applications using a power of 0.5 perform best or as well as the other powers in across all simulation methods. This suggests that the power that leads to a more global explanation is preferred by LIME in this example.

@@

Figures \ref{fig:bullet-eois-nonmatch} and \ref{fig:bullet-eois-match} contain plots that provide this closer look for the known non-match and match, respectively. The top row of plots in each figure are the explanation plots from \emph{lime}. The bottom row of plots show scatterplots of the features selected by LIME colored by the random forest model prediction. The location of the prediction of interest is indicated on the scatterplots by a diamond, and black solid lines are included on the plots representing the bin divisions for applications using bins. The bottom row plots are created using \emph{limeaid}. The plots within a column are associated with one of the three simulation methods (kernel density, 3 equal bins, or 3 quantile bins).

The scatterplots provide a visualization that helps to explain the LIME explanation. For example, consider the explanation for case \Sexpr{bullet_poi_nonmatch} when 3 quantile bins were used. The explanation plot from \emph{lime} shows that case \Sexpr{bullet_poi_nonmatch} having an observed CCF value greater than 0.320 supports a random forest prediction in favor of a match. The scatter plots show that many of the simulated values with CCF greater then 0.320 have random forest predictions greater than 0.5. Additionally, the LIME explanation indicates that a value of matches less than or equal to 1.66 contradicts a prediction in favor of a match. The scatter plots shows that almost all simulated values with a value of matches less than or equal to 1.66 have random forest predictions close to 0. Similar statements can be made about the 3 equal bin explanation for this case and the bin based methods for the known match observation explanations.

@@

\begin{figure*}[!thp]
<<bullet-eois-nonmatch, out.width = '\\textwidth', fig.width = 12, fig.height = 6.25, message = FALSE>>=
# Join the plots
plot_grid(pf2_3qb, pf2_3eb, pf2_kd,
          eoi2_3qb, eoi2_3eb, eoi2_kd,
          nrow = 2,
          rel_heights = c(0.45, 0.55))
@
\caption{Plots of LIME explanations for one case in the test data that is a known non-match. Each column contains plots associated with a different application (3 quantile bins, 3 equal bins, or kernel density). The top row of plots are the explanation plots from \emph{lime}. The bottom row shows scatter plots of the simulated data associated with the prediction of interest. The features plotted are those chosen by LIME in the explanations, and the points are colored by the the random forest predictions. Lines showing the divisions created in the feature space by the bin based LIME methods are included as solid black lines. The scatter plots can be used to better understand the LIME explanation and assess if it is a good explanation.}
\label{fig:bullet-eois-nonmatch}

@@

The scatter plots of the kernel density applications for both the match and non-match case show that almost all of the simulated values have random forest predictions less than 0.5, which provide evidence in favor of a non-match. It appears that simulation method is not providing enough values to cover the feature space that hit regions where the random forest provides predictions above 0.5. The training data has a much smaller amount of matches (\Sexpr{sum(bullet_train$samesource)}) than non-matches (\Sexpr{sum(bullet_train$samesource == FALSE)}), and this simulation method is clearly affected by this imbalance in the two classification categories. Furthermore, this may be the cause of the minimal amount of variability in the LIME explanations across all of the cases in the test data.

Without applying multiple applications LIME to the bullet test data or viewing diagnostic plots of the LIME explanations, it may be very possible to formulate reasons why the LIME explanations make sense. However, the sequence of plots in this section (Figures \ref{fig:bullet-feature-heatmap}, \ref{fig:bullet-lime-expls}, \ref{fig:bullet-assess-metric}, \ref{fig:bullet-eois-match}, and \ref{fig:bullet-eois-nonmatch}) suggest that we should be cautious to trust any of these LIME explanations. While some of them provide an explanation that may be reasonable, there could be better explanations (as seen in the 3 quantile bin case of \autoref{fig:bullet-eois-match}), and some of them provide explanations that make no sense (as seen in \autoref{fig:bullet-lime-expls}). It appears that either LIME needs to be further tuned to provide trustworthy and good explanations, or a different approach may provide better insight. As glimpsed with the scatter plots in Figures \ref{fig:bullet-eois-match} and \ref{fig:bullet-eois-nonmatch}, the more traditional approach to explanation random forest predictions by plotting the random forest predictions over the feature space (partial dependence plots), could provide better insight into the complex model with no need to assess the explanation method if the sampling method is well chosen.

@@

The figures referenced in the previous paragraph from the \data \ example are all based on the application of LIME using a set of tuning parameters. From the use of the feature heatmap, we found that different sets of tuning parameters can produce different explanations. It makes sense that one set of tuning parameters could lead to a more faithful explainer model than another. As a result, we proposed a visual comparison of two faithfulness metrics (MSEE and average fidelity). The examples of faithfulness metric comparisons in this paper (Figures \ref{fig:sine-assess-metric} and \ref{fig:bullet-assess-metric}) both produced confusing results. In both examples, the density based simulation methods resulted in the best or close to the best performance even though the feature heatmaps (Figures \ref{fig:sine-feature-heatmap} and \ref{fig:bullet-feature-heatmap}) showed that the density simulation methods produced global explanations with no variation in features selected by LIME. For the bin based simulation methods, the two metrics often did not agree or contradicted one another, which makes it difficult to decide on a recommendation of a set of tuning parameters that produces the explanations with the most faithful explainer model.
